---
title: "Mixed effects modeling in R"
author: "Paul N. Leiby"
date: "12/29/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

######################################
#                                    #
#   Mixed effects modeling in R     #
#                                    #
######################################

- original authors: Gabriela K Hajduk, based on workshop developed by Liam Bailey
- contact details: gkhajduk.github.io; email: gkhajduk@gmail.com
- date: 2017-03-09 (see [Introduction to linear mixed models](https://ourcodingclub.github.io/2017/03/15/mixed-models.html))

## What is mixed effects modelling and why does it matter?

Ecological and biological data are often complex and messy. We can have different `grouping factors` like populations, species, sites where we collect the data, etc. `Sample sizes` might leave something to be desired too, especially if we are trying to fit complicated models with many parameters. On top of that, our `data points might not be truly independent`. For instance, we might be using quadrats within our sites to collect the data (and so there is structure to our data: quadrats are nested within the sites).

This is why `mixed models` were developed, to deal with such messy data and to allow us to use all our data, even when we have low sample sizes, structured data and many covariates to fit. Oh, and on top of all that, mixed models allow us to save degrees of freedom compared to running standard linear models!

## Explore the data 

We are going to focus on a fictional study system, dragons, so that we don’t have to get too distracted with the specifics of this example. Imagine that we decided to train dragons and so we went out into the mountains and collected data on dragon intelligence (testScore) as a prerequisite. We sampled individuals with a range of body lengths across three sites in eight different mountain ranges. Start by loading the data and having a look at them.

```{r}
### ---- Explore the data -----###
## load the data and have a look at it

load("dragons.RData")

head(dragons)

```

## Let's say we want to know how the body length affects test scores.

## Have a look at the data distribution:

```{r}
hist(dragons$testScore)  # seems close to normal distribution - good!

```

## It is good practice to  standardise your explanatory variables before proceeding - you can use scale() to do that:

```{r}
dragons$bodyLength2 <- scale(dragons$bodyLength)
```

## Back to our question: is test score affected by body length?

###---- Fit all data in one analysis -----###

## One way to analyse this data would be to try fitting a linear model to all our data, ignoring the sites and the mountain ranges for now.

```{r}
library(lme4)

basic.lm <- lm(testScore ~ bodyLength2, data = dragons)

summary(basic.lm)

```

## Let's plot the data with ggplot2

```{r}

library(ggplot2)

ggplot(dragons, aes(x = bodyLength, y = testScore)) +
  geom_point()+
  geom_smooth(method = "lm")

```

### Assumptions?

## Plot the residuals - the red line should be close to being flat, like the dashed grey line

```{r}
plot(basic.lm, which = 1)  # not perfect, but look alright
```

## Have a quick look at the  qqplot too - point should ideally fall onto the diagonal dashed line

```{r}
plot(basic.lm, which = 2)  # a bit off at the extremes, but that's often the case; again doesn't look too bad
```


## However, what about observation independence? Are our data independent?

We collected multiple samples from eight mountain ranges. It's perfectly plausible that the data from within each mountain range are more similar to each other than the data from different mountain ranges - they are correlated. Pseudoreplication isn't our friend.

## Have a look at the data to see if above is true
```{r}
boxplot(testScore ~ mountainRange, data = dragons)  # certainly looks like something is going on here

```

## We could also plot it colouring points by mountain range

```{r}
ggplot(dragons, aes(x = bodyLength, y = testScore, colour = mountainRange))+
  geom_point(size = 2) +
  theme_classic()+
    theme(legend.position = "none")
```

## Plot with points coloured by mountain range, with smoothed linear model for each

```{r}
ggplot(dragons, aes(x = bodyLength, y = testScore, colour = mountainRange))+
  geom_point(size = 2) +
  geom_smooth(method = "lm") 
  theme_classic()+
    theme(legend.position = "none")
```

## From the above plots it looks like our mountain ranges vary both in the dragon body length and in their test scores. This confirms that our observations from within each of the ranges aren't independent. We can't ignore that.

## So what do we do?

###----- Run multiple analyses -----###


## We could run many separate analyses and fit a regression for each of the mountain ranges.

## Lets have a quick look at the data split by mountain range
## We use the facet_wrap to do that

```{r}
ggplot(aes(bodyLength, testScore), data = dragons) + geom_point() +
    facet_wrap(~ mountainRange) +
    xlab("length") + ylab("test score")
```



##----- Modify the model -----###

## We want to use all the data, but account for the data coming from different mountain ranges

## Add mountain range as a fixed effect to our basic.lm

```{r}
mountain.lm <- lm(testScore ~ bodyLength2 + mountainRange, data = dragons)
summary(mountain.lm)
```

## now body length is not significant


###----- Mixed effects models -----###



##----- First mixed model -----##

### model

### plots

### summary

### variance accounted for by mountain ranges



##-- implicit vs explicit nesting --##

head(dragons)  # we have site and mountainRange
str(dragons)  # we took samples from three sites per mountain range and eight mountain ranges in total

### create new "sample" variable


##----- Second mixed model -----##

### model

### summary

### plot



##----- Model selection for the keen -----##

### full model

### reduced model

### comparison
